---
title: "The Effect of Vitamin C on Tooth Growth in Guinea Pigs"
subtitle: "A basic data analysis"
author: "Diego Rabatone Oliveira"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE)
opts_chunk$set(fig.width=4, fig.height=3, warning=FALSE, message=FALSE)
options(digits = 3)
library(tidyr)
library(dplyr)
library(ggplot2)
library(datasets)
```

# Overview
This is the part 2 of the final project for the Statistical Inference Course,
offered by Johns Hopkins University at Coursera, as part of the Data Science
Specialization. This project consists a basic exploratory data analysis of the
ToothGrowth dataset (from R datasets package).

# Loading the data and doing some exploratory analysis
Let's start by loading our dataset:
```{r}
data("ToothGrowth")
TG <- ToothGrowth
glimpse(TG)
```

As we can see, this dataset contains 60 observations with 3 variables (**len** 
(dbl), **sup** (fact: *OJ* and *VC*) and **dose** (dbl)). Looking at the 
`help(ToothGrowth)`, we can see that **dose** is a measure in **mg**, growth
does not present a unit measure, and *OJ* stands for *OrangeJuice* and *VC* for
*Ascorbic Acid*, both being delivery methods from Vitamin C. And all these data
came from a study of the "*Effect of Vitamin C on Tooth Growth in Guinea Pigs*".

# Dataset Summary
Let's now do some summaries from our dataset:
```{r}
summary(TG)
```

The **dose** variable, despite being numeric, only presents the values (0.5, 1 
and 2). Thus, we will convert it to a factor variable in order to threat them as
groups.
```{r}
TG$dose <- as.factor(TG$dose)
```

Now let's do some boxplotting to observe the overall behaviour of these three 
variables togheter, and we are also going to calculate the mean and standard
deviation of the tooth *length*, grouped by *dose* and *supp*, only *dose* and
only *supp*:
```{r}
ggplot(TG, aes(x=factor(dose), y=len)) + 
  facet_grid(.~supp) +
  geom_boxplot(aes(fill = supp)) +
  labs(title="Tooth length by dosage and by type of supplement", 
    x="Dose (mg/day)",
    y="Tooth length")
head(bind_rows(TG %>% group_by(dose, supp) %>% summarize(mean=mean(len), sd=sd(len)),
      TG %>% group_by(dose) %>% summarize(supp=NA, mean=mean(len), sd=sd(len)),
      TG %>% group_by(supp) %>% summarize(dose=NA, mean=mean(len), sd=sd(len))) %>%
  arrange(dose, supp), n=11)
```
From the boxplot above and  the table, it looks like there is a tendency that
the higher the dosage, the larger the tooth length. But let's do some tests to
verify or refute this hypothesis

# Confidence Interval and Hypothesis Testing
Our dataset is not too large, n=`r nrow(TG)`, and when grouping the observations
it will be even smaller (`n/2` on the best case, grouping only by *supp*). So,
our standard errors tends to be bigger than what we would want. Therefore, we
will need to use the `T distribution` for our hypotesis tests. We are assuming
that the given data respects the `Independent and identically distributed random
variables (i.i.d)` criteria, despite no information about this was given.

We are going to start with two tests, and, depending on its results we will do 
another complementary test.

## Supplement types (**supp**)
```{r}
test1 <- t.test(formula=len ~ supp, data=TG, var.equal = FALSE)
```
Here our hypothesis is that there are differences between the mean tooth length
among the groups that received different supplements:
$H_0: \mu_{OJ} \neq \mu_{VC}$. The test 
`t.test(formula=len ~ supp, data=TG, var.equal = FALSE)` results in a `p-value`
of `r test1$p.value`, which is larger than the significance value of 0.05. Being
so, we *fail to reject the null hypothesis* that there are differences on the
means between the *OJ* and the *VC* groups.

## Dosage (**dose**)
```{r}
test2.a <- TG %>% filter(dose==0.5 | dose==1) %>% t.test(formula=len ~ dose, data=., var.equal=FALSE)
test2.b <- TG %>% filter(dose==0.5 | dose==2) %>% t.test(formula=len ~ dose, data=., var.equal=FALSE)
test2.c <- TG %>% filter(dose==1 | dose==2) %>% t.test(formula=len ~ dose, data=., var.equal=FALSE)
ta <- c('u_0.5-u_1=0', test2.a$p.value, test2.a$conf.int[1], test2.a$conf.int[2])
tb <- c('u_0.5-u_2=0', test2.b$p.value, test2.b$conf.int[1], test2.b$conf.int[2])
tc <- c('u_1-u_2=0', test2.c$p.value, test2.c$conf.int[1], test2.c$conf.int[2])
cnames <- c('null hypothesis', 'p.value', 'conf.low', 'conf.high')
result <- data.frame(0,0,0,0)
colnames(result) <- cnames
result[1,] <- ta
result[2,] <- tb
result[3,] <- tc
```
Now, our hypothesis is that there are differences between the mean tooth length 
among the groups that received different dosages, independently of the 
supplement type. Here, as we have three different levels of dosage, we will need
to do three t-tests to cover the possible combinations. The results of these
tests are below:
```{r}
result
```
So, on the three tests done the `p-values` where smalled than the significance
level of 0.05, which means that we **reject the null hypothesis** of not
existing difference between the observations while varying the dosage. So, our
data have strong evidence that the vitamin C dosage influences the average tooth
length on guinea pigs with 95% of confidence.

## Conclusions
Considering the tests done and graphs evaluated, we can conclude that, 
statistically speaking, the Guinea Pigs tooth length presents a strong
relationship with the vitamin C consumption in terms of dose levels, with a 95%
confidence, and does not present relationship with how this vitamin is ingested.

*The source code used to procude this report can be found at:*
https://github.com/diraol/CourseraProjects
\newpage
# Appendix (Codes)
Below are the codes used to buid this report:

```{r, eval=FALSE, echo=TRUE}
options(digits = 3)
library(tidyr)
library(dplyr)
library(ggplot2)
library(datasets)
data("ToothGrowth")
TG <- ToothGrowth
glimpse(TG)
summary(TG)
TG$dose <- as.factor(TG$dose)
ggplot(TG, aes(x=factor(dose), y=len)) + 
  facet_grid(.~supp) +
  geom_boxplot(aes(fill = supp)) +
  labs(title="Tooth length by dosage and by type of supplement", 
    x="Dose (mg/day)",
    y="Tooth length")
head(bind_rows(TG %>% group_by(dose, supp) %>% summarize(mean=mean(len), sd=sd(len)),
      TG %>% group_by(dose) %>% summarize(supp=NA, mean=mean(len), sd=sd(len)),
      TG %>% group_by(supp) %>% summarize(dose=NA, mean=mean(len), sd=sd(len))) %>%
  arrange(dose, supp), n=11)
test1 <- t.test(formula=len ~ supp, data=TG, var.equal = FALSE)
test2.a <- TG %>% filter(dose==0.5 | dose==1) %>% t.test(formula=len ~ dose, data=., var.equal=FALSE)
test2.b <- TG %>% filter(dose==0.5 | dose==2) %>% t.test(formula=len ~ dose, data=., var.equal=FALSE)
test2.c <- TG %>% filter(dose==1 | dose==2) %>% t.test(formula=len ~ dose, data=., var.equal=FALSE)
ta <- c('u_0.5-u_1=0', test2.a$p.value, test2.a$conf.int[1], test2.a$conf.int[2])
tb <- c('u_0.5-u_2=0', test2.b$p.value, test2.b$conf.int[1], test2.b$conf.int[2])
tc <- c('u_1-u_2=0', test2.c$p.value, test2.c$conf.int[1], test2.c$conf.int[2])
cnames <- c('null hypothesis', 'p.value', 'conf.low', 'conf.high')
result <- data.frame(0,0,0,0)
colnames(result) <- cnames
result[1,] <- ta
result[2,] <- tb
result[3,] <- tc
result
```